---
title: "Introduction to humapr"
author: "Benjamin Skov Kaas-Hansen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to humapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = "center",
  fig.height = 4,
  fig.width = 3,
  comment = "#>"
)
```

This vignette gives a quick and dirty introduction to `humapr`. Built on `ggplot2`, it's not all that difficult to use, and we strive to keep the quirks to a minimum. This vignettes features only customisation with standard `ggplot2` functions. 

This first version of humapr is simple: you initiate things with `humap()`, basically creating an empty `ggplot` object with appropriate theme settings, and then you add the `geom` you want to use. We currently have two geom types: a stylised body surface divided into major topographic regions, and stylised internal organs. The arguments to these similar, and we'll show you examples of both in this vignette.

Before we get started, we'll just load useful packages and create a little utility function to help us create simulated data for the examples below.

```{r, message=FALSE}
for (p in c("dplyr", "humapr", "gridExtra", "grid")) 
    library(p, character.only = TRUE)

sampler <- function(x, n) 
    sample(x, n, TRUE, sample(seq(0.5, 1, length = length(x)), length(x), TRUE))
```

## geom_body: Major topographic regions
It shouldn't be a big surprised that this geom maps your data onto the stylised body surface. We start out by creating some simulated data that we can play around with:
```{r}
body_data <- data.frame(loc = sampler(c("head", "neck", "chest", "abdomen", "arm", "forearm", 
                                        "hand", "thigh", "leg", "foot"), 200),
                        side = sampler(c("left", "right"), 200),
                        gender = sample(c("male", "female"), 200, TRUE),
                        year = 2010:2019)
knitr::kable(sample_n(body_data, 10), align = "c", format = "markdown")
```

And now, for a simple plot. We're not giving `humapr` any data to discriminate left- from right-side data, so regions are pooled.
```{r}
humap() +
    geom_body(aes(loc = loc), body_data) 
```

If we do tell `humapr` which variable in the data frame holds information on side of the observation, this is put to use without further arguments:
```{r}
humap() +
    geom_body(aes(loc = loc, side = side), body_data) 
```

## geom_organs: Internal organs
Again, we first create some simulated data to play with.

```{r}
organs <- c("urogenital", "right_kidney", "left_kidney", "pancreas", "liver", "spleen", 
            "oesophagus", "stomach", "intestine", "heart", "trachea", "hyoid", "larynx", 
            "right_lung", "left_lung")
organ_data <- data.frame(loc = sampler(organs, 300),
                         gender = sampler(c("Female", "Male"), 300),
                         period = sampler(c("2001-05", "2006-10"), 300))
```

Then, we make a simple tabulation plot of the observation in each organ:
```{r}
humap() +
    geom_organs(aes(loc = loc), organ_data)
```

## Annotation lines and labels
By default, `humapr` puts annotation lines with absolute and relative frequency labels. You can change this with the `annotate` argument:
```{r, fig.width=4}
humap() +
    geom_body(aes(loc = loc, side = side), body_data, annotate = "all") 
```

```{r}
humap() +
    geom_body(aes(loc = loc), body_data, annotate = NULL) 
```

## Change the fill scale
`humapr` uses the default fill settings of `ggplot`, so that low-count areas are dark, and high-count areas are light. It's straightfrward to change this with the `scale_fill_gradient`. 
```{r}
humap() +
    geom_organs(aes(loc = loc), organ_data) +
    scale_fill_gradient(low = "grey95", high = "darkgreen")
```

## Facetting
Facetting is as simple as always when using `ggplot2`.
```{r, fig.width=6}
humap() +
    geom_organs(aes(loc = loc), organ_data) +
    facet_wrap(~ period)
```

Beware, though, that facetting can give distorted results, because the range is shared across facets. This may be intended behaviour, but sometimes it's better to do "stratified tabulation" and manually align plots. 

To give the same appearance as normal `ggplot` facetting, I do actually use `facet_wrap` even though there is just one facetting level. It's a slight hack, but works well. Keeping annotations here can become a bit unwieldy, so I've removed them in this example.
```{r, fig.width=4}
males <- humap() + 
    geom_organs(aes(loc = loc), filter(organ_data, gender == "Male"), annotate = NULL) +
    facet_wrap(~ gender)
females <- humap() + 
    geom_organs(aes(loc = loc), filter(organ_data, gender == "Female"), annotate = NULL) +
    facet_wrap(~ gender)
grid.arrange(males, females, nrow = 1)
```

